---
title: "RNAseq of the Flor yeast"
author: "Clara Font and Anna Casas"
date: "2026-04-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QUESTION 1: Load, adapt the data and create metadata

### LOADING THE DATA

``` {r}
rep1 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep1.csv"), header=TRUE)
rep2 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep2.csv"), header=TRUE)
rep3 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep3.csv"), header=TRUE)
rep4 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep4.csv"), header=TRUE)

```

### MERGE THE DATASETS
``` {r}
#Merge datasets (without the "X" column so they are not duplicated)
merged_df <- cbind(rep1, rep2[,-1], rep3[,-1], rep4[,-1])

#Define the gene names as row names (I tried without defining and then they are "lost" during transposition)
rownames(merged_df) <- merged_df$X
merged_df$X <- NULL  #Remove the now redundant column

#Transposing (having the dataframe as it was asked in the question 1)
final_df <- as.data.frame(t(merged_df))
#Check dimensions
dim(final_df) #correct, 48 samples as explained in the introduction of the project and +6000 genes

#since PCA works only with numeric data I make sure that the dataframe is prepared for it. 
final_df <- data.frame(lapply(final_df, as.numeric))
rownames(final_df) <- rownames(t(merged_df))


```

After trying different versions, I think I have the dataframe as expected and in a good way to work on it in the next questions and exercises. 

```{r cars}
#str(final_df) (to check that all the variables are numeric)
#View(final_df)
#Columns of the tables have to represent variables (genes) and rows have to be samples.
```

### METADATA
A metadata dataframe explains properties from the original dataframe (what each column and row actually represents). In the case of our dataframe, this information can be extracted from the names of each sample (ex: A.t1_rep1).

```{r}
#get the sample names from the dataframe
sample_names <- rownames(final_df)

#break the names into parts (using strsplit)
metadata <- data.frame(
  Sample = sample_names,  
  SherryWine = substr(sample_names, 1, 1),  #extracting A or B - the first character regarding the type of wine 
  Time = gsub(".*\\.(.*)_.*", "\\1", sample_names), #extracting t1, t2, t3... - between . and _ in the sample name
  Replicate = gsub(".*_", "", sample_names) #extracting replicate number (after _)
) 

#View(metadata)
```

## QUESTION 2: PCA representation

Performing a PCA will allow us to convert a high dimension dataset into a lesser dimension dataset, but still ensuring that the information it provides is similar and useful. 


```{r}
#str(final_df)
head(final_df[, 1:5])

#create the PCA 
pca1 <- stats::prcomp(final_df, scale=TRUE)

pca2<- stats::prcomp(final_df, scale=FALSE)

norm_df <-final_df/rowSums(final_df)
pca3 <- stats::prcomp(norm_df, scale=TRUE)

#plotting the 3 dataframes after the PCA's
library(ggplot2)
library(ggfortify)

ggplot2::autoplot(pca1, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA scaled")

ggplot2::autoplot(pca2, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA non-scaled")

ggplot2::autoplot(pca3, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA normalized")

log_df <- log2(final_df + 1)
pca4 <- stats::prcomp(log_df, scale=TRUE, na.rm=TRUE)
```

The normalized plot looks very strange, with 2 independent clusters and a sample (outlier). We do not really understand why does it look like that. 





