---
title: "RNAseq of the Flor yeast"
author: "Clara Font and Anna Casas"
date: "2026-04-30"
output: 
  html_document: 
    theme: yeti
    highlight: pygments
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, fig.align = "center")
```


```{=html}

<style>
.justify {
  text-align: justify;
}

h1.title {
  color: #F08080;
  font-weight: bold;
  text-align: center;
  margin-bottim: 10px
  letter-spacing: 4px;
}

h4.author, h4.date {
  text-align: center;
  color: #F89880; 
  margin-bottom: 10px;
}

h1, h2, h3, h4, h5, h6 {
  color: #FF8674; 
}

.fig {width: 50%}

.center {
  text-align: center;
}

.alternative_box {
  padding: 1em;
  background: white;
  color: black;
  border: 2px solid darkturquoise;
  border-radius: 10px;
  margin-bottom: 10px;
}

.question_box {
  padding: 1em;
  background: white;
  color: black;
  border: 2px solid plum;
  border-radius: 10px;
  margin-bottom: 10px;
}

</style>
```


:::{.question_box}
## INTRODUCTION

You are a bioinformatician working for a company that provides scientific, statistical and bioinformatic consulting services. A customer from a very important sherry wine trade company wants to **understand the differences between strains of Saccharomyces cerevisiae for different sherry wines products that they elaborate**.

As you may know, flor strains of this **yeast are principal microbial agents responsible for biological wine aging used for production of wines**. The flor yeast velum formed on the surface of fortified fermented must is a major adaptive and technological characteristic of flor yeasts that helps them to withstanding stressful winemaking conditions and ensures specific biochemical and sensory oxidative alterations typical for sherry wines.
The user produces two types of sherry wines (A and B), with different flavors, alcohol percentage and color. They are interested in deciphering if the differences between sherry wines classes are due to genetic differences in yeast or any other biochemical process that has not been controlled.
We recommended to the user applying **RNAseq technology for transcriptome analysis of these two industrial flor yeasts strains at different steps of velum development (6 different times) and multiple replicates (4 replicates)**. We ended up with 48 samples analyzed.

RNAseq reads were automatically quality checked, trimmed and filtered following default settings of packages such as fastqc and trimmomatic included in a RNA seq bioinformatics pipeline (https://github.com/HCGB-IGTP/RNA_seq_pipe). During the quality check, a sample produced unusual results but was not discarded. Please consider this and discard it if necessary after the normalization process. 

Mapping to the reference genome and feature count of genes was produced using Hisat2 and featureCounts packages, also included in the previous pipeline mentioned. For each of the replicates, a gene count matrix was generated and saved as a comma separated file within the data folder provided. Metadata can be extracted from the names of the samples.

Your **goal** in this project is to decipher by applying dimension reduction techniques if there are genetic differences between sherry wine yeast strains that might be reflected in the wine tasting.
:::


## QUESTION 1: Load, adapt the data and create metadata

### LOADING THE DATA

``` {r}
rep1 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep1.csv"), header=TRUE)
rep2 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep2.csv"), header=TRUE)
rep3 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep3.csv"), header=TRUE)
rep4 <- read.csv(file.path("/home/clarafontc/homeassignment/", "rep4.csv"), header=TRUE)

```

### MERGE THE DATASETS
``` {r}
#Merge datasets (without the "X" column so they are not duplicated) 
#(we apply directly cbind since we have seen that samples appear in the same order in all the datasets)
merged_df1 <- cbind(rep1, rep2[,-1], rep3[,-1], rep4[,-1])

#Define the gene names as row names (I tried without defining and then they are "lost" during transposition)
rownames(merged_df1) <- merged_df1$X
merged_df1$X <- NULL  #Remove the now redundant column

#Transposing (having the dataframe as it was asked in the question 1)
final_df1 <- as.data.frame(t(merged_df1))
#Check dimensions
dim(final_df1) #correct, 48 samples as explained in the introduction of the project and +6000 genes

#since PCA works only with numeric data I make sure that the dataframe is prepared for it. 
final_df1 <- data.frame(lapply(final_df1, as.numeric))
rownames(final_df1) <- rownames(t(merged_df1))


```

After trying different versions, I think I have the dataframe as expected and in a good way to work on it in the next questions and exercises. 

```{r cars}
#str(final_df1) (to check that all the variables are numeric)
#View(final_df1)
#Columns of the tables have to represent variables (genes) and rows have to be samples.
```

### METADATA
A metadata dataframe explains properties from the original dataframe (what each column and row actually represents). In the case of our dataframe, this information can be extracted from the names of each sample (ex: A.t1_rep1).

```{r}
#get the sample names from the dataframe
sample_names <- rownames(final_df1)

#break the names into parts (using strsplit)
metadata <- data.frame(
  Sample = sample_names,  
  SherryWine = substr(sample_names, 1, 1),  #extracting A or B - the first character regarding the type of wine 
  Time = gsub(".*\\.(.*)_.*", "\\1", sample_names), #extracting t1, t2, t3... - between . and _ in the sample name
  Replicate = gsub(".*_", "", sample_names) #extracting replicate number (after _)
) 

#View(metadata)
```

## QUESTION 2: PCA representation

### DOING THE PCA'S

Performing a PCA will allow us to convert a high dimension dataset into a lesser dimension dataset, but still ensuring that the information it provides is similar and useful. 


```{r, warning=FALSE, message=FALSE}
#str(final_df)
head(final_df1[, 1:5])

#create the PCA 
pca1 <- stats::prcomp(final_df1, scale=TRUE)

pca2<- stats::prcomp(final_df1, scale=FALSE)

norm_df <-final_df1/rowSums(final_df1)
pca3 <- stats::prcomp(norm_df, scale=TRUE)

#plotting the 3 dataframes after the PCA's
library(ggplot2)
library(ggfortify)

ggplot2::autoplot(pca1, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA scaled") 

ggplot2::autoplot(pca2, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA non-scaled")

plotnorm <- ggplot2::autoplot(pca3, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA normalized")

library(plotly)
ggplotly(plotnorm)

autoplot(pca3, data=metadata, label=TRUE)

```

The normalized plot looks very strange, with 2 independent clusters and a sample (outlier). Possible batch effect in the data will be discussed later.
By using the "autoplot" line (using label=TRUE) I have been able to easily identify which is the outlier in the previous plot (the normalized one). The assignment says "A sample produced unusual results but was not discarded.", so I guess it is our choice whether to discard it or not. We will see this sample with more detail to make a decision. 

### CHECKING THE OUTLIER
```{r}
#View(rep3)

total_counts <- colSums(merged_df1[,-1])

barplot(
  total_counts,
  las = 2,
  main = "Checking the outlier",
  ylab = "Total counts"
)

```

This simple graph helps to understand the dimension of the outlier and its difference with the rest of values. That sample strongly separates from all others and doesn't cluster with the replicates (seen in PCA graph of normalized data). This can be due to a technical issue, perhaps a sequencing error, a mapping failure,... but we discard that it might be due to biological factors. That's why we conclude that the optimal choice is to leave Repetition 3 of the first time from experiment B (B.t1_rep3) out of the study. 


```{r}
#new dataframes without the outlier

merged_df2 <- merged_df1[, colnames(merged_df1) != "B.t1_rep3"]
final_df2 <- final_df1[rownames(final_df1) != "B.t1_rep3", ]
norm_df2 <- norm_df[rownames(norm_df) != "B.t1_rep3", ]
metadata <- metadata[metadata$Sample!="B.t1_rep3", ]

#performing the normalized pca again
pca4 <- stats::prcomp(norm_df2, scale=TRUE)

ggplot2::autoplot(pca4, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA - normalized data - by Wine Type")
```

We can see how the data appears less clustered since the PCA has worked properly with the normalized data. Removing the outlier has been a correct choice. 

### BATCH EFFECT

The PCA graphs have been colored by wine type and batch effect will now be evaluated To keep on studying the batch effect we will plot everything but coloring it by other parameters (Replicates and Time) and using PCA with the new dataframes (without the outlier). We will use scale=TRUE since we think is the best for this kind of studies.

```{r}
#performing the graphs 

ggplot2::autoplot(pca4, data = metadata, colour = 'Replicate') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA - normalized data - by Replicate")

ggplot2::autoplot(pca4, data = metadata, colour = 'Time') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA - normalized data - by Time")

```

Starting by the graph colored by Wine Type, we can see that PC1 captures most of the variance in your dataset (interpreting the x-axis). Data is splitted along the x axis (negative part, Wine A and positive part Wine B). This implies that the most important difference in gene expression across the samples is given by the differences in strains of yeast between the two wine kinds (2 different strains of Saccharomyces cerevisiae for 2 different sherry wine products they offer), not due to a batch effect. 

Following by the graph colored by Replicate, it shows different replicates overlapping as different subgroups. This means that batch effect doesn't cause the variance, so there have not been significant technical problems while conducting the transcriptome analysis. 

The last graph to see carefully is the one colored by Time. We can see that the samples are distributed by color along the y-axis (PC2), which explains a small percentage of the variance in the data. Taking into account that flor yeast velum is adaptative and changes over time (wine aging), it makes sense that samples have variance along the different times they have been analyzed. 

### CORRELATION OF AXIS WITH EXPRESSION

```{r}

#creating the dataframe of the pca and the metadata to plot and see correlations. 
pca_scores <- as.data.frame(pca4$x)
pca_scores <- cbind(pca_scores, metadata)

#plotting using PC1 (the one that explains most of the variance)
ggplot(pca_scores, aes(x = SherryWine, y = PC1, color = SherryWine)) +
  geom_jitter() +
  theme_minimal() +
  labs(title = "PC1 - Wine Type")

ggplot(pca_scores, aes(x = Time, y = PC1, color=SherryWine)) +
  geom_jitter() +
  theme_minimal() +
  labs(title = "PC1 - Time")

#After trying many graphs (using PC1, PC2 and even PC3, I think the one that shows a correlation as important as to appear here is PC2 with Time).
ggplot(pca_scores, aes(x = Time, y = PC2, color = SherryWine)) +
  geom_jitter() +
  theme_minimal()+
  labs(title="PC2 - Time")

ggplot(pca_scores, aes(x = Replicate, y = PC1, color = SherryWine)) +
  geom_jitter() +
  theme_minimal()+
  labs(title="PC1 - Replicate")

```

PC1 explains most of the variance (65.1%) and as we can see it clearly shows separation between the 2 types (A and B), which reafirms that the transcriptomic differences between strains of Saccharomyces cerevisiae are the origin of the variation. 

The variance explained by PC1 when it comes to Time is clearly shown in the graph, where samples appear differently distributed along the y-axis (PC1) depending on the Wine Type they belong to. PC2 (which explains a much smaller part of the variance, a 6.98%) shows a weak negative correlation with the time the samples were taken. 

Finally, the graph between PC1 and Replicates looks pretty similar to the one between this PC and Time. Samples differ on the y-axis depending on the Wine Type. This explains once again that most of the variance is explained by the biological difference between the 2 flor yeast strains rather than the effects Time or Replicates might have. No clusters are shown, which once again confirms that there is no batch effect. 


## QUESTION 3: tSNE representation

### REPRESENT THE DATA WITH A TSNE

tSNE is a technique used to reduce dimensionality of data. PCA is used with the same goal, but it has some cons, such as projecting the data focusing on keeping separated the distant points, and not taking into account that it should keep close points together. Also it is not suitable for non-linear data. 
tSNE is a non-linear feature extraction, it is an algorithm that tries to embed the points from a higher dimension to a lower dimension trying to preserve the neighborhood of that point (from the slides). 

We will now apply it to our datasets, raw data and normalized data. We will set the perplexity parameter (says how to balance attention between local and global aspects of your data) to 15 since the default value is 30 and it does not work with the dimension of the datasets. 

To allow the plots to be reproducible we will set a seed when calling the Rtsne function (in this case to 42).

```{r}
set.seed(42)
library("Rtsne")
tsne1 <-Rtsne(final_df2, perplexity=15)
tsne2 <- Rtsne(norm_df2, perplexity=15)
```

To interpret the tSNE's properly, we will plot the data. To make the plots more easilly I have decided to create dataframes with the data necessary for the plots (everything tidier and more accessible).

```{r}
plotdata1 <- data.frame(
  tSNE1 = tsne1$Y[, 1],          
  tSNE2 = tsne1$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time, 
  Replicate = metadata$Replicate)

ggplot(plotdata1, aes(x = tSNE1, y = tSNE2, color = Replicate)) +
  geom_point() + 
    labs(title = "tSNE plot - raw data by Replicate",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

ggplot(plotdata1, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - raw data by Wine Type",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 


plotdata2 <- data.frame(
  tSNE1 = tsne2$Y[, 1],          
  tSNE2 = tsne2$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time,
  Replicate = metadata$Replicate)

ggplot(plotdata2, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - normalized data",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

```

After trying all possible graphs to see whether there might be a batch effect or not, we have decided to show these ones. 

The first graph corresponding to the Raw Data colored by the Replicate shows 4 separate regions based on the Replicate number. This might indicate sequencing differences. To confirm if this is a batch effect we use the normalized data, since it is not technically biased. With this method we can conclude that the batch effect does not show anymore (samples are separated uniquely by Wine Type, reaffirming a possible biological difference among the 2 flor yeast strains). 

### CORRELATION OF AXIS WITH EXPRESSION

To follow with the study, we will do as in the PCA dimensionality reduction, we will try to identify if there is an axis correlation with expression. This time we will do it in a different way, since tSNE is a non-linear method, and the axes do not explain the percentage of variance (as in the PCA). We will be analyzing the possible correlation from the graphs (normalized data). Since the normalized data graph is pretty simple, with the Wine Type A samples at the top left part of the grid and Wine Type B diagonally placed (bottom right), we cannot say it is strongly correlated with the axis (tSNE1 or tSNE2) as we saw in the PCA since there are just 2 groups of samples depending on the Wine Type. What we can actually say is that the tSNE is strongly associated with the WineType (categorical variable), as it shows by the differenciated groups in the graph. 
Once again, the clusters shown may not actually explain a linear correlation but they do show transcriptomic differences between strains of Saccharomyces cerevisiae.

## QUESTION 4: tSNE parameters

As we did in the tSNE practical, we will now test how different values for Perplexity, Iterations and Reproducibility work on the dimensionality reduction of the data using the tSNE method. 

If we check help("Rtsne"), we can see the default values. They are 30 for Perplexity (as mentioned before), 1000 for Iterations and Reproducibility is a changes every single time (tSNE is stochastic, it requires a random number to start the analysis). 

### PERPLEXITY

I will start exploring the parameter "perplexity" and using plots to see the differences more clearly. 

```{r}

tsne3 <- Rtsne(norm_df2, perplexity=1)
tsne4 <- Rtsne(norm_df2, perplexity=15)

plotdata3 <- data.frame(
  tSNE1 = tsne3$Y[, 1],        
  tSNE2 = tsne3$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time, 
  Replicate = metadata$Replicate)

ggplot(plotdata3, aes(x = tSNE1, y = tSNE2, color = Time)) +
  geom_point() + 
    labs(title = "tSNE plot - perplexity = 1",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

plotdata4 <- data.frame(
  tSNE1 = tsne4$Y[, 1],          
  tSNE2 = tsne4$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time,
  Replicate = metadata$Replicate)

ggplot(plotdata4, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - perplexity=15",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 


```

The lowest value I use is 1 and the highest is 15, since for this dataframe the tSNE can't be performed for higher values. 

For the perplexity = 1 plot, we can see that the values appear much very distributed along the grid, clustered in groups of 4 (after checking every part of the metadata table, I have seen that the subgroups depend on the Time variable).
For perplexity = 15, it is clear that we have 2 determined groups in diagonal depending on the Wine Type. Data is not distributed, there are 2 groups. 

I have also checked for perplexity = 7, which is an intermediate value. The distribution of the samples looks much more similar to perplexity = 1 than 15. Data is distributed but grouped by Time. When trying values of perplexity like 9 or higher, the samples start being separated to the diagonals depending on they Type of strains. 

### ITERATIONS

We will keep on by exploring different values of max_iter. 

```{r}
tsne5 <- Rtsne(norm_df2, perplexity=15, max_iter=10)
tsne6 <- Rtsne(norm_df2, perplexity=15, max_iter=100)
tsne7 <- Rtsne(norm_df2, perplexity=15, max_iter=1000)

plotdata5 <- data.frame(
  tSNE1 = tsne5$Y[, 1],        
  tSNE2 = tsne5$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time, 
  Replicate = metadata$Replicate)

ggplot(plotdata5, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - iterations = 10",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

plotdata6 <- data.frame(
  tSNE1 = tsne6$Y[, 1],          
  tSNE2 = tsne6$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time,
  Replicate = metadata$Replicate)

ggplot(plotdata6, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - iterations = 100",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

plotdata7 <- data.frame(
  tSNE1 = tsne7$Y[, 1],          
  tSNE2 = tsne7$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time,
  Replicate = metadata$Replicate)

ggplot(plotdata7, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - iterations = 1000",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

```

I have decided to check the iteration values 10, 100 and 1000. 
It can be easily seen from the graphs how the more iterations the less overlapping there is between samples of different Wine Type. The more iterations, the more clustered the data appears. 

### REPRODUCIBILITY 
The last parameter left to explore is reproducibility:

```{r}
set.seed(10)
tsne8 <- Rtsne(norm_df2, perplexity=15)
set.seed(100)
tsne9 <- Rtsne(norm_df2, perplexity=15)
set.seed(1000)
tsne10 <- Rtsne(norm_df2, perplexity=15)

plotdata8 <- data.frame(
  tSNE1 = tsne8$Y[, 1],        
  tSNE2 = tsne8$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time, 
  Replicate = metadata$Replicate)

ggplot(plotdata8, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - seed = 10",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

plotdata9 <- data.frame(
  tSNE1 = tsne9$Y[, 1],          
  tSNE2 = tsne9$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time,
  Replicate = metadata$Replicate)

ggplot(plotdata9, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - seed = 100",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

plotdata10 <- data.frame(
  tSNE1 = tsne10$Y[, 1],          
  tSNE2 = tsne10$Y[, 2],          
  WineType = metadata$SherryWine,  
  Time = metadata$Time,
  Replicate = metadata$Replicate)

ggplot(plotdata10, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - seed = 1000",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

```
To see how the reproducibility affects the tSNE's, I have used set.seed(), to choose which value did I want and have it set so that it did not change every time me or anyone else ran it (have the same starting point). As it appears in the graphs, there is no pattern for reproducibility, in contrast to perplexity or iterations. It changes everytime, given that tSNE is not a deterministic algorithm (such as PCA). It follows a random probability distribution (stochastic), so there is a part that is not under "control". 

## QUESTION 5: Final interpretation

### FINAL REPRESENTATIONS

I will start by giving the plot that from our point of view better represents the PCA, which is the one made from normalized data without the outlier and colored by Wine Type (it helps explain the genetic differences between sherry wine yeast strains). 

```{r}
ggplot2::autoplot(pca4, data = metadata, colour = 'SherryWine') +
  ggplot2::theme_minimal() +
  ggplot2::ggtitle("PCA - normalized data - by Wine Type")

```

For the tSNE, the best plot I think that is the one from normalized data and colored by Wine Type as well. We will set perplexity=15 since it is the most optimal in this case. 

```{r}
ggplot(plotdata2, aes(x = tSNE1, y = tSNE2, color = WineType)) +
  geom_point() + 
    labs(title = "tSNE plot - normalized data",
       x = "tSNE1",
       y = "tSNE2") +  theme_minimal() 

```

In our opinion, these graphs are the ones that better represent the whole project (clean data without the outlier, PCA from normalized data and tSNE with a proper Perplexity value, showing the differences between flor yeast strains, which was the whole goal of the assignment). 

### CONCLUSIONS PC1, PC2

After performing the PCA without the outlier and with the normalized data, we have identified that most of the variance of the dataset is explained by PC1 (as it appears in the graph), and seggregates the 2 Wine Kinds, which supports the hypothesis of the genetic differences between the 2 yeast strains being the reason of the taste differences between Wine Types. 
When it comes to PC2, it explains a much smaller percentage of the variance but it has a correlation with the variable Time. It explains how the flor yeast velum adapts to the biological aging process of wine, as it was explained in the presentation of the dataset. 


### CONCLUSIONS

After starting to plot the dataset, we identified an outlier and valued the decision to remove it or not, which turned out to be a yes. There was a significant difference between that sample and the others, and we thought that the best was to take it apart from the study. This methods have provided us proofs of what we expected, that there are biological/genetic differences between the strains of Saccharomyces cerevisiae that provide those different tastes in the Wine produced. These microbial agents that help provoke wine aging conduct the biological differences between the 2 Wine Types (A and B), and this is confirmed by the PCA and the tSNE (dimension reduction techniques). The normalization process has helped to better understand the data without any possible bias. 
